---
title: "198: What a win costs"
date: "2026-09-29"
categories:
  - gtUtils
  - flexoki
  - cfbplotR
  - ggthemes
format:
  html:
    code-fold: true
    code-summary: "Code"
editor: visual
---

Three charts on the same spine: how many points it takes to win an FBS game,
what one real week of that looked like, and one team that hasn't been able to
get there.

All three read from local snapshots of collegefootballdata.com taken on
September 28, 2026, so the post rebuilds without hitting the API.

```{r setup, warning=FALSE, message=FALSE}

# One family across every chart here. Installed at ~/Library/Fonts; swap to
# "sans" if this is ever rebuilt somewhere it isn't available.
chart_font <- "Roboto Condensed"

if (!any(grepl(chart_font, systemfonts::system_fonts()$family, fixed = TRUE))) {
  stop(chart_font, " not found. Restart R after installing, then re-run.")
}

panel_bg <- "#F0F0F0"
bar_outline <- "#5a5a5a"
reference_line <- "#2b2b2b"
carolina_blue <- "#56a0d3"
loss_pink <- "lightpink"

# Five-point bands, shared by the chart below and the analysis behind it.
bin_levels <- c(
  "0-4", "5-9", "10-14", "15-19", "20-24", "25-29", "30-34", "35-39", "40+"
)
```

### Chart: what a win costs

```{r win-cost-chart, warning=FALSE, message=FALSE, fig.width=10, fig.height=6}

# Every team-game from 2024 onward, bucketed by that team's own final score,
# with the bar being the share of those team-games that were wins. So the
# chart reads as a price list: score this much, win this often.
#
# Note the denominator. Each game contributes two rows -- the winner's score
# and the loser's. That's what makes a win *rate* possible; a chart of
# winning scores alone has no denominator to divide by, and every bar would
# sit at 100%.

by_bin <- readr::read_csv("win_cost_by_bin_2024_2026.csv", show_col_types = FALSE) |>
  dplyr::mutate(bin = factor(bin, levels = bin_levels)) |>
  dplyr::arrange(bin) |>
  dplyr::mutate(
    favoured = win_pct >= 50,
    # Sample size rides in the axis label rather than inside the bar: the
    # 0-4 and 5-9 bars are too short to hold text without colliding with
    # their own percentage.
    bin_label = paste0(bin, "\n", format(team_games, big.mark = ","))
  ) |>
  # As a bare character the axis would sort alphabetically and put 10-14
  # ahead of 5-9, so pin the order to the bins themselves.
  dplyr::mutate(bin_label = factor(bin_label, levels = bin_label))

# The first band that wins more often than it loses.
crossover <- by_bin |> dplyr::filter(favoured) |> dplyr::slice_head(n = 1)

# The rate chart gets its own ramp rather than the blue/pink used on the
# North Carolina chart further down. Blue and pink mean "win" and "loss" for
# a specific team there; reusing them here, where every bar is a rate rather
# than an outcome, would read as the same encoding when it isn't. The ramp is
# keyed to the rate itself, so colour and bar height reinforce each other.
#
# Palette is flexoki's green, a single-hue ramp running pale olive to
# near-black. Only the fill comes from flexoki -- the theme stays
# theme_fivethirtyeight() so this matches the rest of the set.
cost_accent <- unname(flexoki::flexoki$green[["green700"]])

# "Team-game" is the honest unit but it's jargon, so the subtitle spells the
# counting out instead: every game is read twice, once from each sideline,
# and each side is filed under whatever that team scored.
subtitle <- sprintf(
  paste0(
    "Every FBS game from 2024-2026, once from each team's sideline, counted ",
    "under what that team scored. For example, a 31-24 game counts once at 31 ",
    "and once at 24, %s entries in all. Each bar is how often teams scoring ",
    "that much went on to win. Above the line, a win is more likely than a ",
    "loss; that starts at <b style='color:%s;'>%s points</b>. FBS vs FBS ",
    "games only, FCS opponents excluded."
  ),
  format(sum(by_bin$team_games), big.mark = ","), cost_accent, crossover$bin[1]
)

cost_chart <- ggplot2::ggplot(
  by_bin, ggplot2::aes(x = bin_label, y = win_pct, fill = win_pct)
) +
  ggplot2::geom_col(width = 0.76, color = bar_outline, linewidth = 0.2) +
  ggplot2::geom_hline(
    yintercept = 50, color = reference_line, linetype = "dashed", linewidth = 0.5
  ) +
  ggplot2::geom_text(
    ggplot2::aes(label = sprintf("%.0f%%", win_pct), y = win_pct + 3.4),
    family = chart_font, fontface = "bold", size = 4.6, color = "#2b2b2b"
  ) +
  ggplot2::annotate(
    "text", x = 0.62, y = 54, hjust = 0,
    label = "Wins more often than it loses",
    family = chart_font, fontface = "bold", size = 3.6, color = reference_line
  ) +
  flexoki::scale_fill_flexoki_c(palette = "green", guide = "none") +
  ggplot2::scale_y_continuous(
    limits = c(0, 104), breaks = seq(0, 100, 25),
    labels = function(x) paste0(x, "%"),
    expand = ggplot2::expansion(mult = c(0, 0.02))
  ) +
  ggplot2::labs(
    title = "What a Win Costs",
    subtitle = subtitle,
    x = "Points a team scored (how many times that happened)",
    y = "Share of those games won",
    caption = paste0(
      "Viz by Chris at Bless your Chart<br>data via collegefootballdata.com, ",
      "regular season, FBS vs. FBS"
    )
  ) +
  ggthemes::theme_fivethirtyeight(base_family = chart_font) +
  ggplot2::theme(
    legend.position = "none",
    plot.title = ggplot2::element_text(
      face = "bold", size = 22, margin = ggplot2::margin(b = 10)
    ),
    plot.subtitle = ggtext::element_textbox_simple(
      size = 11.5, color = "#4a4a4a", margin = ggplot2::margin(b = 16), lineheight = 1.3
    ),
    # element_markdown, not element_text: the caption carries a <br>, which a
    # plain text element prints literally instead of breaking the line.
    plot.caption = ggtext::element_markdown(
      size = 9, color = "#5a5a5a", margin = ggplot2::margin(t = 12), lineheight = 1.3
    ),
    axis.title = ggplot2::element_text(size = 10.5, color = "#4a4a4a"),
    axis.title.x = ggplot2::element_text(margin = ggplot2::margin(t = 6)),
    axis.title.y = ggplot2::element_text(margin = ggplot2::margin(r = 6)),
    axis.text = ggplot2::element_text(size = 11, color = "#3c3c3c"),
    axis.text.x = ggplot2::element_text(lineheight = 1.25),
    panel.grid.major.x = ggplot2::element_blank(),
    panel.grid.minor = ggplot2::element_blank()
  )

ggplot2::ggsave(
  "win_cost_distribution.png", cost_chart,
  width = 10, height = 6, dpi = 300, bg = panel_bg
)

cost_chart
```

### GT table: what a win cost in week four

```{r week-four-table, warning=FALSE, message=FALSE}

# One real week next to the distribution above, so the spread of winning
# scores can be seen rather than taken on faith.
#
# Scope is P4 vs. P4 (ACC, Big 12, Big Ten, SEC) plus any Notre Dame game --
# Notre Dame is independent but plays a P4-caliber schedule, so its games
# count. The winning score carries the same green ramp as the chart above, so
# a high-scoring win looks the same in both places. The losing side is
# deliberately inert: grey score, desaturated logo.

# Light-to-dark green, taken from flexoki. Stops short of the near-black end
# of that ramp: these are table cells with text sitting on them, not bars.
green_ramp <- c("#DDE2B2", "#BEC97E", "#A0AF54", "#768D21", "#536907")

team_logo_url <- function(team) {
  cfbplotR::logo_ref$logo[
    match(cfbplotR::clean_school_names(team), cfbplotR::logo_ref$school)
  ]
}

logo_img <- function(team, grey = FALSE, height = 28) {
  style <- paste0(
    "height: ", height, "px; width: auto; vertical-align: -30%;",
    if (isTRUE(grey)) " filter: grayscale(100%) opacity(0.45);" else ""
  )
  glue::glue("<img src='{team_logo_url(team)}' style='{style}'>")
}

wk4_raw <- readr::read_csv("week_scores_2026_wk4.csv", show_col_types = FALSE)

week_num <- 4

date_span <- range(wk4_raw$game_date)
span_label <- if (date_span[1] == date_span[2]) {
  format(date_span[1], "%B %e, %Y")
} else {
  paste0(format(date_span[1], "%B %e"), "–", format(date_span[2], "%e, %Y"))
}

wk4 <- wk4_raw |>
  dplyr::arrange(dplyr::desc(win_score), dplyr::desc(win_score - lose_score)) |>
  dplyr::transmute(
    winner = paste0(
      "<span style='font-weight:700;'>", win_team, "</span>&nbsp;&nbsp;",
      logo_img(win_team)
    ),
    win_score,
    sep = "–",
    lose_score,
    loser = paste0(
      logo_img(lose_team, grey = TRUE),
      "&nbsp;&nbsp;<span style='color:#8a8a8a;'>", lose_team, "</span>"
    )
  )

wk4_gt <- wk4 |>
  gt::gt() |>
  gtUtils::gt_theme_savant() |>
  gt::fmt_markdown(columns = c(winner, loser)) |>
  gt::data_color(
    columns = win_score,
    palette = green_ramp,
    # Flips the number to light type once the cell goes dark.
    autocolor_text = TRUE
  ) |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold", size = gt::px(22)),
    locations = gt::cells_body(columns = win_score)
  ) |>
  gt::tab_style(
    style = gt::cell_text(color = "#9a9a9a", size = gt::px(20)),
    locations = gt::cells_body(columns = lose_score)
  ) |>
  gt::tab_style(
    style = gt::cell_text(color = "#bdbdbd", size = gt::px(16)),
    locations = gt::cells_body(columns = sep)
  ) |>
  gt::cols_align(align = "right", columns = winner) |>
  gt::cols_align(align = "center", columns = c(win_score, sep, lose_score)) |>
  gt::cols_align(align = "left", columns = loser) |>
  gt::cols_label(
    winner = "Winner", win_score = "", sep = "", lose_score = "", loser = "Loser"
  ) |>
  # Kept tight so the two scores read as one scoreline rather than two
  # separate columns; the winning cell still needs enough width to carry its
  # green fill as a block.
  gt::cols_width(
    winner ~ gt::px(215),
    win_score ~ gt::px(46),
    sep ~ gt::px(14),
    lose_score ~ gt::px(34),
    loser ~ gt::px(215)
  ) |>
  gt::tab_header(
    title = gt::html(paste0("<b>What a Win Cost in Week ", week_num, "</b>")),
    subtitle = paste0(
      "Every P4-vs-P4 result plus Notre Dame, ", span_label,
      ", sorted by winning score"
    )
  ) |>
  gt::tab_source_note(gt::html(paste0(
    "<hr>Winning score is shaded on a green gradient: darker is a higher total. ",
    "P4 is the ACC, Big 12, Big Ten and SEC; Notre Dame is independent, but included.<br>",
    "<hr>Table by Chris at Bless your Chart | data via collegefootballdata.com | cfbtxt.com"
  ))) |>
  gt::tab_options(
    table.font.size = gt::px(15),
    data_row.padding = gt::px(5),
    heading.align = "left"
  )

gtUtils::gt_save_crop(
  wk4_gt,
  file = "week_scores_2026_wk4.png",
  zoom = 2,
  whitespace = 40
)

wk4_gt
```

### Chart: North Carolina against the threshold

```{r unc-streak-chart, warning=FALSE, message=FALSE, fig.width=14, fig.height=12}

# North Carolina's points scored in every FBS game since 2024, against the two
# numbers that matter: 30 points (the streak threshold) and the national
# average winning score.
#
# Postseason is included, which is load-bearing here -- the streak's first
# game is the 2024 Fenway Bowl. Drop bowls and it reads 13 instead of 14.

streak_color <- "#c0392b"

# Kickoff timestamps are stored in UTC, so five of North Carolina's night
# games land on the following calendar day unless converted -- the 2024
# Minnesota opener is stored as 2024-08-30T00:00Z but was a Thursday 8pm ET
# kickoff on 8/29. Leading zeros are stripped by hand because the "%-m" glibc
# extension isn't available on macOS.
format_game_date <- function(ts) {
  paste0(
    as.integer(format(ts, "%m", tz = "America/New_York")), "/",
    as.integer(format(ts, "%d", tz = "America/New_York")), "/",
    format(ts, "%Y", tz = "America/New_York")
  )
}

# 2025-26 only, and reconciles across sources: bcftoys' game logs give 34.83
# for this window, collegefootballdata.com gives 34.70.
national_avg_win <- readr::read_csv(
  "national_avg_winning_score.csv", show_col_types = FALSE
)$avg_winning_score[1]

unc_all <- readr::read_csv(
  "unc_fbs_games_2024_2026.csv",
  col_types = readr::cols(kickoff_utc = readr::col_character())
) |>
  dplyr::mutate(
    kickoff = as.POSIXct(kickoff_utc, format = "%Y-%m-%dT%H:%M:%SZ", tz = "UTC")
  ) |>
  dplyr::arrange(kickoff) |>
  dplyr::mutate(
    game_idx = dplyr::row_number(),
    won = unc_pts > opp_pts,
    hit30 = unc_pts >= 30,
    logo_team = cfbplotR::clean_school_names(opponent),
    result_label = paste0(
      dplyr::if_else(won, "W", "L"), ", ", unc_pts, "-", opp_pts
    ),
    # Date and location carry the left axis; the result rides out past the
    # opponent logo, so each row reads date -> bar -> logo -> outcome.
    x_label = paste0(format_game_date(kickoff), "\n", location)
  )

# The streak: every game since the last 30-point game, which is the 2024
# regular-season finale vs. NC State (exactly 30 in a 35-30 loss).
last30_idx <- max(unc_all$game_idx[unc_all$hit30])
streak <- unc_all |> dplyr::filter(game_idx > last30_idx)
streak_len <- nrow(streak)

# The 4-9 record and 17.3 scoring average belong to the Belichick era (2025
# onward), which is NOT the same set of games as the streak. The streak
# reaches one game further back, to the 2024 Fenway Bowl, so its own record is
# 4-10 and its own average 17.1. The annotation says "under Belichick"
# explicitly so the two windows don't get conflated.
belichick <- unc_all |> dplyr::filter(season >= 2025)
bb_record <- sprintf("%d-%d", sum(belichick$won), sum(!belichick$won))
bb_ppg <- mean(belichick$unc_pts)
# Happens to equal the streak high too, since the only pre-2025 game in the
# streak is the 14-point Fenway Bowl.
bb_max <- max(belichick$unc_pts)

unc_chart <- ggplot2::ggplot(unc_all, ggplot2::aes(x = game_idx, y = unc_pts)) +
  ggplot2::annotate(
    "rect", xmin = last30_idx + 0.5, xmax = max(unc_all$game_idx) + 0.5,
    ymin = 0, ymax = Inf, fill = streak_color, alpha = 0.08
  ) +
  ggplot2::geom_hline(
    yintercept = national_avg_win, color = reference_line,
    linetype = "dashed", linewidth = 0.5
  ) +
  ggplot2::geom_hline(
    yintercept = 30, color = "#7a7a7a", linetype = "dotted", linewidth = 0.5
  ) +
  # A thin outline keeps the pale bars legible against the light panel.
  ggplot2::geom_col(
    ggplot2::aes(fill = won), width = 0.72, color = bar_outline, linewidth = 0.2
  ) +
  cfbplotR::geom_cfb_logos(
    ggplot2::aes(x = game_idx, y = unc_pts + 2.9, team = logo_team),
    width = 0.032, inherit.aes = FALSE
  ) +
  ggplot2::geom_text(
    ggplot2::aes(label = unc_pts), hjust = 1.4, color = "#1a1a1a",
    fontface = "bold", size = 3.7, family = chart_font
  ) +
  # Two passes. The tinted label is 30% opaque, so on rows whose score sits
  # near a reference line the dashed rule shows straight through the box and
  # looks like the text is clipped (the Syracuse 27-10 row is the bad one).
  # This opaque underlay in the panel colour masks the line first; the tinted
  # label then draws on top and keeps its win/loss wash.
  ggplot2::geom_label(
    ggplot2::aes(y = unc_pts + 5.2, label = result_label),
    hjust = 0, color = NA, fill = panel_bg, size = 3.2,
    family = chart_font, linewidth = 0
  ) +
  ggplot2::geom_label(
    ggplot2::aes(y = unc_pts + 5.2, label = result_label, fill = won),
    hjust = 0, color = "#333333", size = 3.2, fontface = "bold",
    family = chart_font, alpha = 0.3, linewidth = 0.25
  ) +
  # Sits at the foot of the streak band rather than the bottom of the chart:
  # high enough to be seen, and on a row whose bar and result label both stop
  # well short of the dashed line, so nothing collides with it.
  ggplot2::annotate(
    "text", x = last30_idx + 1.6, y = national_avg_win + 1.2, hjust = 0,
    label = sprintf("National average winning score (%.1f)", national_avg_win),
    color = reference_line, size = 3.9, fontface = "bold", family = chart_font
  ) +
  ggplot2::annotate(
    "text", x = max(unc_all$game_idx) - 0.5, y = 66, hjust = 1,
    label = sprintf(
      "%d: most points scored against\nan FBS opponent under Belichick", bb_max
    ),
    color = streak_color, size = 5.0, fontface = "bold",
    family = chart_font, lineheight = 1.1
  ) +
  ggplot2::annotate(
    "text", x = max(unc_all$game_idx) - 3.4, y = 66, hjust = 1,
    label = sprintf(
      "Under Belichick: %s vs FBS\n%.1f points per game", bb_record, bb_ppg
    ),
    color = "#4a4a4a", size = 4.75, fontface = "bold",
    family = chart_font, lineheight = 1.2
  ) +
  ggplot2::scale_fill_manual(
    values = c("TRUE" = carolina_blue, "FALSE" = loss_pink)
  ) +
  # Headroom past 50 so the result labels clear the widest bar and its logo.
  ggplot2::scale_y_continuous(
    limits = c(0, 67), breaks = seq(0, 50, 10),
    expand = ggplot2::expansion(mult = c(0, 0.02))
  ) +
  # Not reversed: under coord_flip the highest game index lands at the top,
  # putting the most recent game first and running backwards in time.
  ggplot2::scale_x_continuous(
    breaks = unc_all$game_idx,
    labels = unc_all$x_label,
    expand = ggplot2::expansion(add = 0.8)
  ) +
  ggplot2::coord_flip(clip = "off") +
  ggplot2::labs(
    title = sprintf(
      paste0(
        "North Carolina hasn't scored 30 or more points against an FBS ",
        "opponent in %d consecutive games"
      ),
      streak_len
    ),
    # The subtitle colours the words "win" and "loss" to match the bars, so
    # the legend doesn't have to be spelled out.
    subtitle = sprintf(
      paste0(
        "Carolina's points scored in every FBS game since 2024: a ",
        "<b style='color:%s;'>win</b> or a <b style='color:%s;'>loss</b>; ",
        "dotted line is 30 points, dashed line the national average winning score."
      ),
      carolina_blue, loss_pink
    ),
    caption = paste0(
      "Viz by Chris at Bless your Chart<br>data via collegefootballdata.com | ",
      "FBS opponents only; includes the 2024 Fenway Bowl."
    )
  ) +
  ggthemes::theme_fivethirtyeight(base_family = chart_font) +
  ggplot2::theme(
    legend.position = "none",
    plot.title = ggplot2::element_text(
      face = "bold", size = 17, margin = ggplot2::margin(b = 4)
    ),
    plot.subtitle = ggtext::element_markdown(
      size = 12, color = "#4a4a4a", margin = ggplot2::margin(b = 14)
    ),
    plot.caption = ggtext::element_markdown(
      size = 9, color = "#5a5a5a", margin = ggplot2::margin(t = 12), lineheight = 1.3
    ),
    # The date/location stack is supporting detail, so it sits smaller than
    # the annotations it's read alongside.
    axis.text.y = ggplot2::element_text(
      size = 7.2, color = "#4a4a4a", hjust = 1, lineheight = 1.15
    ),
    axis.text.x = ggplot2::element_text(size = 10.5, color = "#3c3c3c"),
    panel.grid.major.y = ggplot2::element_blank(),
    panel.grid.major.x = ggplot2::element_line(color = "#cccccc", linewidth = 0.3),
    panel.grid.minor = ggplot2::element_blank()
  )

ggplot2::ggsave(
  "unc_points_vs_threshold.png", unc_chart,
  width = 14, height = 12, dpi = 300, bg = panel_bg
)

unc_chart
```
